---
title: "R Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyr)
```

#Reads csv file
```{r}
csv_path <- "C:/Users/danas/Documents/School/Earth Systems R/data/"
csv_name <- "20190724_Georeferenced_ABBRV_2018AnnualSAR_0.csv"
csv_file <- paste(csv_path, csv_name, sep="")
sar_csv <- read.csv(csv_file)
```

```{r}
names(sar_csv)
unique(sar_csv$State)
```


```{r}
missions_by_state <- sar_csv %>%
  group_by(State) %>%
  summarise(
    total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) +sum(Unfounded),
    total_fatalities = sum(Fatality),
    total_injuries = sum(Ill_or_Injured),
    total_non_fatalities = sum(Not_Ill_or_Injured),
    total_unfound = sum(Unfounded),
    total_air_operations = sum(NPS_Air_Operations_Costs != 0) + sum(Military_Air_Operations_Costs != 0)
  ) %>%
  arrange(desc(total_missions))

missions_by_park <- sar_csv %>%
  group_by(Park) %>%
  summarise(
    total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) +sum(Unfounded),
    total_fatalities = sum(Fatality),
    total_injuries = sum(Ill_or_Injured),
    total_non_fatalities = sum(Not_Ill_or_Injured),
    total_unfound = sum(Unfounded),
    total_air_operation = sum(NPS_Air_Operations_Costs != 0) + sum(Military_Air_Operations_Costs != 0)
  ) %>%
  arrange(desc(total_missions))

head(missions_by_state)
head(missions_by_park)

missions_by_state <- missions_by_state %>% arrange(desc(total_fatalities))
print(missions_by_state)


missions_by_park <- missions_by_park %>% arrange(desc(total_fatalities))
print(missions_by_park)

missions_by_state <- missions_by_state %>% arrange(desc(total_unfound))
print(missions_by_state)

missions_by_park <- missions_by_park %>% arrange(desc(total_unfound))
print(missions_by_park)



```
Top Three States for Missions
California : 468 Missions
Utah : 411 Missions
Arizona : 356 Missions

Top Three National Parks for Missions: 
Grand Canyon (AZ) : 333
Sequoia and Kings Canyon (CA) : 237
Rocky Mountain (CO) : 217

Top Three States with Highest Recovery Missions (Fatal):
Arizona : 23
California : 20
Washington : 12

Top Three National Parks with Highest Recovery Missions:
Organ Pipe Cactus (AZ) : 13
Golden Gate (CA) : 10
Grand Canyon (AZ) : 9

Top Three National Parks with Most Unfound People
Glacier Bay (AK)
Grand Teton (WY)
Pinnacles (CA) : 8




#Index plot of SAR incidents lat/long
```{r}
plot(sar_csv$LONGITUDE, sar_csv$LATITUDE, 
     xlab = "Longitude", ylab = "Latitude",
     main = "Scatter Plot of Longitude vs. Latitude")
```
```{r}
# Create a data frame with columns Fatality, Male, and Female
df <- data.frame(sar_csv$Fatality, sar_csv$Male, sar_csv$Female)

# Rename the columns for clarity
colnames(df) <- c("Fatality", "Male", "Female")

# Load the required library
library(ggplot2)
library(tidyr)

# Reshape the data from wide to long format
df_long <- pivot_longer(df, cols = c("Male", "Female"), names_to = "Gender", values_to = "Count")

# Create the bar plot
ggplot(df_long, aes(x = Fatality, y = Count, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Fatality Comparison by Gender",
       x = "Fatality", y = "Count") +
  scale_fill_manual(values = c("Male" = "skyblue", "Female" = "pink")) +
  theme_minimal()

```

Index plot of lat long SAR incidents
```{r}
plot(sar_csv$LONGITUDE, sar_csv$LATITUDE, 
     xlab = "Longitude", ylab = "Latitude",
     main = "Scatter Plot of Longitude vs. Latitude")
```


```{r}


state_summary <- sar_csv %>%
  group_by(State) %>%
  summarise(total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) + sum(Unfounded))

# Convert State to factor with levels ordered by total_missions
state_summary$State <- factor(state_summary$State, levels = state_summary$State[order(state_summary$total_missions, decreasing = TRUE)])


# Create the bar plot
ggplot(state_summary, aes(x = State, y = total_missions)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "State Comparison by Total Missions",
       x = "State", y = "Total Missions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




California Fatality by State Park
```{r}
# Create a data frame with columns Park and Fatality

ca_data <- subset(sar_csv, State == "CA")
df <- data.frame(Park = ca_data$Park, Fatality = ca_data$Fatality)


# Reshape the data from wide to long format
df_long <- pivot_longer(df, cols = "Fatality", names_to = "Fatalities", values_to = "Count")

ggplot(df_long, aes(x = Park, y = Count, fill = Fatalities)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "California Parks Comparison by Fatality",
       x = "Park", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
California Park Comparasion by Mission 
```{r}
ca_data <- subset(sar_csv, State == "CA")

# Summarize data by park
park_summary <- ca_data %>%
  group_by(Park) %>%
  summarise(total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) + sum(Unfounded ))


# Create the bar plot
ggplot(park_summary, aes(x = Park, y = total_missions)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = " California Parks Comparison by Total Missions",
       x = "Park", y = "Total Missions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
az_data <- subset(sar_csv, State == "AZ")

# Summarize data by park
park_summary <- az_data %>%
  group_by(Park) %>%
  summarise(total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) + sum(Unfounded ))


# Create the bar plot
ggplot(park_summary, aes(x = Park, y = total_missions)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = " Arizona Parks Comparison by Total Missions",
       x = "Park", y = "Total Missions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



Arizona Fatality by State Park
```{r}
# Create a data frame with columns Park and Fatality

az_data <- subset(sar_csv, State == "AZ")
df <- data.frame(Park = az_data$Park, Fatality = az_data$Fatality)


# Reshape the data from wide to long format
df_long <- pivot_longer(df, cols = "Fatality", names_to = "Fatalities", values_to = "Count")

# Create the bar plot
ggplot(df_long, aes(x = Park, y = Count, fill = Fatalities)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = " Arizona Parks Comparison by Fatality",
       x = "Park", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
ut_data <- subset(sar_csv, State == "UT")

# Summarize data by park
park_summary <- ut_data %>%
  group_by(Park) %>%
  summarise(total_missions = sum(Ill_or_Injured) + sum(Not_Ill_or_Injured) + sum(Unfounded ))


# Create the bar plot
ggplot(park_summary, aes(x = Park, y = total_missions)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Utah Parks Comparison by Total Missions",
       x = "Park", y = "Total Missions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



Utah Fatality by State Park
```{r}
# Create a data frame with columns Park and Fatality

ut_data <- subset(sar_csv, State == "UT")
df <- data.frame(Park = ut_data$Park, Fatality = ut_data$Fatality)


# Reshape the data from wide to long format
df_long <- pivot_longer(df, cols = "Fatality", names_to = "Fatalities", values_to = "Count")

# Create the bar plot
ggplot(df_long, aes(x = Park, y = Count, fill = Fatalities)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Utah Parks Comparison by Fatality",
       x = "Park", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Comparing Recovery Missions (Deceased Subject) vs Rescues (Live Subject)
```{r}

us_states_sf <- st_read("C:/Users/danas/Documents/School/Earth Systems R/data/cb_2018_us_state_500k.shp")

# Subset data based on fatalities
subset_data <- subset(sar_csv, !is.na(Fatality))

# Plot
ggplot() + 
  geom_sf(data = us_states_sf, fill = NA) + 
  geom_point(data = subset_data, aes(x = LONGITUDE, y = LATITUDE, size = Fatality, color = ifelse(Fatality == 0, "Non-Fatal", "Fatal"))) +
  scale_color_manual(name = "Fatality",
                     values = c("Non-Fatal" = "blue", "Fatal" = "red"),
                     labels = c("Fatal Mission", "Non_Fatal Mission")) +
  scale_size_continuous(name = "Number of Fatalities") +
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50), expand = FALSE) +  
  theme_bw()
             
```



Missions Injured vs Non Injured Subjects

```{r}
library(dplyr)


# Filter data where injury occurred
injuries_true <- sar_csv[sar_csv$Ill_or_Injured == TRUE, ]

# Filter data where no injury occurred
injuries_false <- sar_csv[sar_csv$Ill_or_Injured == FALSE, ]

# Plot
ggplot() + 
  geom_sf(data = us_states_sf, fill = NA) +
  geom_point(data = injuries_true, aes(x = LONGITUDE, y = LATITUDE), color = "red") +
  geom_point(data = injuries_false, aes(x = LONGITUDE, y = LATITUDE), color = "blue") +
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50), expand = FALSE) +
  theme_bw()


```


Missions Where Subject Was Not Located
```{r}
unfounded_true <- sar_csv[sar_csv$Unfounded == TRUE, ]



# Plot
ggplot() + 
  geom_sf(data = us_states_sf, fill = NA) +
  geom_point(data = unfounded_true, aes(x = LONGITUDE, y = LATITUDE), color = "red") +
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50), expand = FALSE) +
  theme_bw()
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

